---
title: "STAT656_LendingClubPJ"
author: "Pei-Yin Yang, Monica Daniel, Sebastian Bravo Sanchez"
date: "2/17/2022"
output: html_document
---
### Reading in Data and Loaging Libraries
#### Loading Libraries

```{r importlibs}
packs = c('caret','corrplot', 'e1071','readr', 'tidyverse','RANN',
          'AppliedPredictiveModeling','stringr')
lapply(packs,require,character.only=TRUE)
```

#### Loading Data*

*I removed rows 39788 and 39789 from the LoanStats3a.csv file because they were blank except for the indication the the subsequent rows were for loans granted out of policy. I also created a new column called 'meets_policy' that takes value of 'Yes' for the loans that meet the policy (first 39786) and 'No' for the ones that don't (comment by Sebastian).

```{r readdata}
LCData <-  read.csv('LoanStatsEdited.csv',header=TRUE,stringsAsFactors = FALSE,
                    na.strings=c("","NA"))
is.data.frame(LCData)
LCDataTB <-  as_tibble(LCData) #convert it to tibble
set.seed(1)
rowsMissingStatus <-  -which(LCDataTB$loan_status == "")
#LCDataTB <- LCDataTB %>% slice(rowsMissingStatus)

#created if statement because there were zero missing rows after I edited the csv file
if (length(rowsMissingStatus)>0){
  LCDataTB <- LCDataTB %>% slice(-which(LCDataTB$loan_status == ""))
}
```

### Preprocessing

#### Checking loan status types

```{r}
#Loan Statuses(Supervisor) check
LCDataTB %>% group_by(loan_status) %>% summarise(n = n())
anyNA(LCDataTB$loan_status)
```

#### Cleaning and subsetting supervisor (Loan Status)

```{r}
#Rename loan status rows and remove rows where loan status is not of interest
LCDataTB <-  LCDataTB %>% 
    mutate(loan_status = replace(loan_status, loan_status == "Does not meet the credit policy. Status:Charged Off", "Charged Off")) %>% 
    mutate(loan_status = replace(loan_status, loan_status == "Does not meet the credit policy. Status:Fully Paid", "Fully Paid")) %>% 
    filter(loan_status == 'Charged Off' | loan_status == 'Fully Paid') 
LCDataTB %>% group_by(loan_status) %>% summarise(n = n())
```

#### Checking variable types and content

```{r}
# Checking data types
table(sapply(LCDataTB[1,], class))
# Checking if there are integers classified as numeric
str(LCDataTB)
```

#### Deleting Unecessary Variables

Deletion based on the following:
 - a) Features with greater than 50% missing data
 - b) Categorical features with just one category
 - c) Business knowledge from analyzing the Data Dictionary. 
 
 Note: We can discuss a) and c) (by Sebastian)

```{r}
# remove features that contain more than 50% missing values
LCDataTB <- LCDataTB[, !sapply(LCDataTB, function(x) mean(is.na(x))) > 0.5]

# Deleting meaningless features that refer to the individual and personal information that is restricted to avoid discrimination:
# Sex, race and ethnicity, zip code (proxy for classes of people), etc.
#TODO: addr_state
# addr_state may not influence the outcome interest? (Pei)
# zip_code maybe a proxy of status, if using for any type of decision making, can lead to a serious discrimination suit (Pei)

#LCDataTB <- select(LCDataTB, -c(id,member_id,url,desc,zip_code,addr_state))

## Try not to remove id##
LCDataTB <- select(LCDataTB, -c(member_id,url,desc,zip_code,addr_state))

# Deleting pricing-related information already contained in other variables: the installment amount is already contained in the debt-to-income (DTI). The interest rate is needed for calculating the installment amount, but it is a result of the credit scoring, not a predictor.
LCDataTB <- select(LCDataTB, -c(int_rate,installment))

# Remove total_acc & grade
LCDataTB <- select(LCDataTB, -c(total_acc,grade))


# remove categorical features with only one category
LCDataTB <-  LCDataTB %>% select(where(~n_distinct(.) > 1)) #(Sebastian)

# TODO:Remove features that contain one unique value + NA
names(LCDataTB[, sapply(LCDataTB, function(col) length(unique(col))) == 2])
## Remove "collections_12_mths_ex_med" & "chargeoff_within_12_mths" because they have only one unique value and NA
LCDataTB <- LCDataTB %>%
  select(-c(collections_12_mths_ex_med,chargeoff_within_12_mths))

```

#### Mutating Non-uniform Categorical Features

The objective is to check some of the fields with open text ('emp_title' and 'title') and summarizing them into something more compact. Another option is to do some text mining and discover more information from the text, but I suggest we look into this only if we have time after producing a minimum viable product. (comment by Sebastian)

```{r}
#Checking unique values
(unique = LCDataTB %>% select(emp_title, title) %>% summarise_all(n_distinct))
#Showing the number of replication for each unique value, ordered in descending order
LCDataTB %>% group_by(emp_title) %>% summarise(n = n()) %>% arrange()
LCDataTB %>% group_by(title) %>% summarise(n = n()) %>% arrange()
#Deleting both features
LCDataTB <- LCDataTB %>% select(-c(emp_title,title))
```
'emp_title' and 'title' have `r unique$emp_title` and `r unique$title` unique values, respectively. Therefore, there are very few replications for each value. After confirming the frequency of unique values is very low, it makes sense to delete these features for now.

#### Dictionary Check for feature selection

Pei:
sub_grade is contained in grade; 
open_acc is contained in total_acc

Should we remove from one of each?

```{r}
#Dictionary <- read_excel("LCDataDictionary.xlsx")
#Dictionary <- Dictionary %>% filter(LoanStatNew %in%  colnames(LCDataTB))
#write_csv(Dictionary, file = "Dictionary_check.csv")

#Deleting information not known at the time of origination for credit decision purposes (we can revisit this later)

#TODO: check the dictionary
LCDataTB <- select(LCDataTB, -c(funded_amnt, funded_amnt_inv, total_pymnt, 
                                total_pymnt_inv, total_rec_int, 
                                total_rec_late_fee, recoveries, 
                                collection_recovery_fee, 
                                last_pymnt_d, last_pymnt_amnt,last_credit_pull_d, 
                                total_rec_prncp, issue_d))
```


#### Checking variable types and content II

Now that we reduced the number features, it is easier to check that numeric and categorical features are appropriately classified

```{r}
# Checking data types
table(sapply(LCDataTB[1,],class))
# Checking if there are integers classified as numeric
str(LCDataTB)
# Checking the number of unique values in each column
sapply(LCDataTB, function(col) length(unique(col)))
```

#### Converting to the appropriate data type

The data dictionary helped identify misclassified data types.
The following integer features should be numeric: loan_amt, revol_bal, total_acc

The following character features should be numeric: revol_util, earliest_cr_line (this is a date)

The following integer features should be factors: These are up for debate because most are count data (e.g., delinq_2yrs). 

I think we can encode anything that has 4 or more unique counts as numeric and anything else as factors. (comment by Sebastian).

All other character features would be converted to factors. 

```{r}
LCDataTB %>% select(delinq_2yrs,inq_last_6mths,
                    pub_rec,delinq_amnt,
                    pub_rec_bankruptcies, tax_liens) %>%
    summarise_all(n_distinct)
#TODO: Convert features to the appropriate categories
#TODO: Carve-out and encode the supervisor
#TODO: Impute missing values
#TODO: Split features into factors and numeric in order to apply transformations (center and scale, Box-Cox, etc.) to numeric features

## Step 1: Convert chr and integer with less than 4 levels into factor
Chr_fac <- c("emp_length", "home_ownership",
             "loan_status","purpose","pymnt_plan",
             "sub_grade","term","verification_status", "meets_policy")
Int_fac <- c("acc_now_delinq", "tax_liens")
variableFactor <- c(Chr_fac, Int_fac)

LCDataFactor <- LCDataTB %>%
  select(any_of(variableFactor), earliest_cr_line, id) %>% 
  mutate_all(as.factor)

##############################Investigation purpose, can be removed, Don't run these
# Work on earliest_cr_line (04/13/2022)
#(1)Format: Month-Year: Jan-00
L1 <- LCDataTB %>%
    filter(str_detect(earliest_cr_line, c("^J|^F|^M|^A|^S|^O|^N|^D")))

#(2)Format: Year-Month: 1-Jan
L2 <- LCDataTB %>%
    filter(str_detect(earliest_cr_line, 
                      c("^1|^2|^3|^4|^5|^6|^7|^8|^9|^10|^11|^12|^13|^14|^15|^16|^17|^18|^19|^20")))
#Since # of obs for L1+L2 is not equal to # of obs. for LCDataTB, investigate unmatched records
Lcombine <- rbind.data.frame(L1,L2)
Dif <- anti_join(LCDataTB, Lcombine, by = 'id') #turns out earliest_cr_line in 29 unmatched records are blank
#############################Investigation purpose, can be removed, Don't run these

# Work on earliest_cr_line (04/13/2022)
## Real Conversion starts from here
#(1)Format: Month-Year: Jan-00
L1_factor <- LCDataFactor %>%
    filter(str_detect(earliest_cr_line, c("^J|^F|^M|^A|^S|^O|^N|^D")))

#(2)Format: Year-Month: 1-Jan
L2_factor <- LCDataFactor %>%
    filter(str_detect(earliest_cr_line, 
                      c("^1|^2|^3|^4|^5|^6|^7|^8|^9|^10|^11|^12|^13|^14|^15|^16|^17|^18|^19|^20")))

# Convert the format of earliest_cr_line to Date
L1_factor$earliest_cr_line <- as.Date(sub("$","-01",L1_factor$earliest_cr_line), "%b-%y-%d") 
L2_factor$earliest_cr_line <- as.Date(sub("^","01-",L2_factor$earliest_cr_line), "%d-%y-%b")
Lcombine_factor <- bind_rows(L1_factor,L2_factor)
Diff <- anti_join(LCDataFactor, Lcombine_factor, by = 'id')
Diff$earliest_cr_line <- as.Date(Diff$earliest_cr_line)

LCDataFactor1 <- bind_rows(L1_factor,L2_factor,Diff)

## Step 2: Make a new object for now that just has the quantitative features, converting integer type to numeric

LCDataQuant = LCDataTB %>%
  mutate(revol_util = parse_number(revol_util)/100) %>% #convert character of percentage into numeric
  select(-any_of(variableFactor), -earliest_cr_line) %>% 
  mutate_all(as.numeric)

## Combine the above two steps
LCDataFactor1$id <- as.numeric(levels(LCDataFactor1$id))[LCDataFactor1$id] #convert a factor to integer\numeric without loss of information
LCDataTB <- left_join(LCDataQuant,LCDataFactor1, by = 'id')


# Remove features that have close to zero variance
# Use it once we confirm the variable type
LCDataTB_2 <- LCDataTB[, -nearZeroVar(LCDataTB)] #(Pei)
```

### Preprocessing I (could be replaced by PCA?)

I'm not sure this code chunk should be applied to the entire dataset. In some examples I saw Darren first split categorical and numerical variables before applying pre-processing since not everything applies uniformly, e.g., it does not make sense to center / scale factors nor apply 'nzv'. (comment by Sebastian)

```{r}
# Variability check
# remove features that have minor variability
LCDataTB_2 <- LCDataTB_2 %>% preProcess(method = 'nzv') %>%
    predict(newdata = LCDataTB_2)

#Center/scale for quantitative variables
LCDataTB_2 <- LCDataTB_2 %>% preProcess(.) %>%
    predict(newdata = LCDataTB_2)
```

### PCA analysis

```{r}

```

### Missing data and Imputation

```{r}
#feature-wise deletion
#remove features that contain more than 50% missing values
LCDataTB <- LCDataTB[,!sapply(LCDataTB, function(x) mean(is.na(x))) > 0.5]

#Amputation


```

### Train and Test Split

```{r}
#Xtrain = select(LCDataTB, -loan_status)
#Xtest = select(LCDataTB, -loan_status)
#Ytrain = select(LCDataTB, loan_status) %>% unlist()
#Ytest = select(LCDataTB, loan_status) %>% unlist()
#table(Ytrain)

```
